---
title: "Assignment 5: Data Visualization"
author: "Yufan Du"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
# Load necessary libraries
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)

# Verify home directory
here::here()

# Set the file paths for the datasets
peter_data <- read_csv(here("Data","Processed","Processed_KEY", "NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"))
niwo_data <- read_csv(here("Data","Processed","Processed_KEY", "NEON_NIWO_Litter_mass_trap_Processed.csv"))

#2 
# Check the structure of the data to see how the date columns are being read
str(peter_data)
str(niwo_data)

```

## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
# Load necessary libraries
library(ggplot2)

# Define a custom theme
my_custom_theme <- theme(
  # Customize plot background
  plot.background = element_rect(fill = "lightblue", color = NA), 
  
  # Customize plot title
  plot.title = element_text(size = 12, face = "bold", hjust = 0.5, color = "darkblue"),
  
  # Customize axis labels
  axis.title = element_text(size = 10, color = "darkblue"),
  
  # Customize axis ticks and gridlines
  axis.text = element_text(size = 8, color = "black"),
  axis.ticks = element_line(color = "black"),
  panel.grid.major = element_line(color = "gray80", linewidth = 0.5),
  panel.grid.minor = element_line(color = "gray90", linewidth = 0.25),
  
  # Customize legend
  legend.background = element_rect(fill = "white", color = "black"),
  legend.title = element_text(face = "bold"),
  legend.text = element_text(size = 12)
)

# Set the custom theme as the default
theme_set(theme_bw() + my_custom_theme)


```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
#decide the extreme values to hide
summary(peter_data$po4)
summary(peter_data$tp_ug)

#plot the graph
ggplot(peter_data, aes(x = po4, y = tp_ug, color = lakename)) +
  geom_point(alpha = 0.6, size = 2) +  # Add scatter plot points with transparency
  geom_smooth(method = "lm", se = FALSE) +  # Add linear regression line (line of best fit)
  scale_color_manual(values = c("Peter Lake" = "black", "Paul Lake" = "blue")) +  # Customize colors for lakes
  labs(title = "Total Phosphorus vs Phosphate in Peter and Paul Lakes",
       x = "Phosphate (ug/L)",
       y = "Total Phosphorus (ug/L)",
       color = "Lake Name") +
  xlim(0, 20) +  # Adjust x-axis limits to hide extreme values
  ylim(0, 50) +  # Adjust y-axis limits to hide extreme values
  theme(legend.position = "bottom")  # Move legend to bottom

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

# Ensure the 'month' column is treated as a factor with numeric labels for proper ordering
peter_data$month <- factor(peter_data$month, levels = 1:12, labels = 1:12)

# Plot 1: Temperature boxplot
plot_temp <- ggplot(peter_data, aes(x = month, y = temperature_C, fill = lakename)) +
  geom_boxplot() +
  labs(title = "Temperature by Month", y = "Temperature (°C)") +
  theme(legend.position = "none", axis.title.x = element_blank())  # Remove x-axis title

# Plot 2: Total Phosphorus (TP) boxplot 
plot_tp <- ggplot(peter_data, aes(x = month, y = tp_ug, fill = lakename)) +
  geom_boxplot() +
  labs(title = "Total Phosphorus by Month", y = "Total Phosphorus (ug/L)") +
  theme(legend.position = "none", axis.title.x = element_blank())  # Remove x-axis title

# Plot 3: Total Nitrogen (TN) boxplot 
plot_tn <- ggplot(peter_data, aes(x = month, y = tn_ug, fill = lakename)) +
  geom_boxplot() +
  labs(title = "Total Nitrogen by Month", x = "Month", y = "Total Nitrogen (ug/L)") +
  theme(legend.position = "none")  # Remove legend

# Extract the legend from one of the plots
legend_plot <- get_legend(
  ggplot(peter_data, aes(x = month, y = temperature_C, fill = lakename)) +
    geom_boxplot() +
    theme(legend.position = "bottom")  # Place legend at the bottom
)

# Combine the three plots into one cowplot, ensuring axes are aligned
combined_plots <- plot_grid(
  plot_temp, plot_tp, plot_tn,
  ncol = 1,  # Arrange plots vertically
  align = "v"  # Align the y-axis
)

# Add the shared legend at the bottom
final_plot <- plot_grid(
  combined_plots, legend_plot,
  ncol = 1,  # Place legend below the plots
  rel_heights = c(3, 0.2)  # Adjust the size ratio between the plots and the legend
)

# Display the final combined plot
print(final_plot)



```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Temperature: Shows a clear seasonal pattern, with higher values in warmer months (May–August) and lower in colder months. Both lakes exhibit similar temperature trends.
Total Phosphorus (TP): TP levels vary across months without a strong seasonal pattern. There are notable differences between the lakes, with one lake having higher TP levels at certain times.
Total Nitrogen (TN): TN shows some seasonal variation, with higher levels in late spring/summer. Differences between lakes suggest varying nutrient dynamics.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
# Filter the dataset for "Needles" functional group
needles_data <- niwo_data %>% filter(functionalGroup == "Needles")

# Plot: Dry mass of needle litter by date, colored by NLCD class
ggplot(needles_data, aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  geom_line() +
  labs(title = "Dry Mass of Needle Litter Over Time by NLCD Class",
       x = "Collection Date",
       y = "Dry Mass (g)") +
  theme(legend.position = "bottom")

#7
# Plot: Dry mass of needle litter by date, faceted by NLCD class
ggplot(needles_data, aes(x = collectDate, y = dryMass)) +
  geom_line() +
  facet_wrap(~nlcdClass, ncol = 3) +
  labs(title = "Dry Mass of Needle Litter Over Time, Faceted by NLCD Class",
       x = "Collection Date",
       y = "Dry Mass (g)") +
  theme(legend.position = "bottom")


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: 6. Becuase it's easier to compare in a same graph.